Quantifying the importance of an inference model in Bayesian phylogenetics

### 2020-01-28 content.tex
Call a site model a site model, call a nucleotide substitution model as such, fix #84

The first step simulates a DNA alignment from a given 
phylogeny (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 2a)
using the DNA alignment parameters.
The DNA alignment parameters consist of 
a (DNA) root sequence, a (DNA) mutation rate, a clock model 
and a nucleotide substitution model.
The root sequence is the DNA sequence of the shared common ancestor,
and is set to four different equally-sized mononucleotide blocks by default, as this
helps interpreting the resulting alignment.
Supported nucleotide substitution model, which
we will refer to as site models, 
are JC, HKY, TN and GTR. Only the strict
clock model is currently supported in this step.

@@

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny, as described in Subsection~\ref{subsec:pipeline}.
In this example, the root sequence consists of four blocks of 250 
mononucleotides each, while the per-nucleotide mutation rate is 
0.1 mutations per unit time.
We use a Jukes-Cantor (JC, \cite{jukes1969evolution}) site model
and a strict clock model as these are the simplest.
A JC site model assumes that mutation rates between nucleotides are equal and 
constant. 
A strict clock model assumes that the mutation rates 
of all lineages are equal and constant.

@@

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \texttt{model\_type} &
    \texttt{run\_if} &
    \texttt{do\_measure\_evidence} & 
    \texttt{inference model} \\ 
    \hline
    generative &
    always &
    FALSE &
    JC, strict, Yule \\
    \hline
  \end{tabular}
  \caption{
    Inference conditions and model.
    JC: Jukes-Cantor site model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior.
  }
  \label{tab:RQ1}
\end{table}

@@

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \textbf{model\_type} &
    \textbf{run\_if} &
    \textbf{do\_measure\_evidence} & 
    \textbf{inference model} \\ 
    \hline
    generative & always         & FALSE & JC, strict, Yule \\
    candidate  & best candidate & TRUE  & JC, strict, BD   \\
    candidate  & best candidate & TRUE  & JC, strict, CBS  \\
    ...        & ...            & ...   & ...              \\
    candidate  & best candidate & TRUE  & GTR, RLN, CCP    \\
    candidate  & best candidate & TRUE  & GTR, RLN, CEP    \\
    \hline
  \end{tabular}
  \caption{
    Inference conditions and model.
    JC: Jukes-Cantor site model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior.
    BD: birth-death tree prior.
    GTR: GTR site model.
    RLN: relaxed log-normal clock model.
    CBS: coalescent Bayesian Skyline tree prior.
    CCP: coalescent constant-population tree prior.
    CEP: coalescent exponential-population tree prior.
  }
  \label{tab:RQ2}
\end{table}

@@

Here we specify a different set of experiments: we
need to state that we already have an experiment for
the generative model, as well as that we want all the other inference models 
to compete. We call the competing models 'candidate models'.
As model selection is commonly performed on the full list of available 
candidate models, \verb;pirouette; has a dedicated function 
for this choice: \verb;create_all_experiments; creates a full set 
of 40 experiments, 
containing the inference models of all combinations of 4 site models, 
2 clock models and 5 tree priors. All we need to add is to exclude the 
inference model in the generative experiment:

>>>>

The first step simulates a DNA alignment from a given 
phylogeny (Fig.~\ref{fig:pipeline}, 1a $\rightarrow$ 2a)
using the DNA alignment parameters.
The DNA alignment parameters consist of 
a (DNA) root sequence, a (DNA) mutation rate, a clock model 
and a nucleotide substitution model.
The root sequence is the DNA sequence of the shared common ancestor,
and is set to four different equally-sized mononucleotide blocks by default, as this
helps interpreting the resulting alignment.
Supported nucleotide substitution model
are JC, HKY, TN and GTR. Only the strict
clock model is currently supported in this step.

@@

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny, as described in Subsection~\ref{subsec:pipeline}.
In this example, the root sequence consists of four blocks of 250 
mononucleotides each, while the per-nucleotide mutation rate is 
0.1 mutations per unit time.
We use a Jukes-Cantor (JC, \cite{jukes1969evolution}) nucleotide substitution model
and a strict clock model as these are the simplest.
A JC model assumes that mutation rates between nucleotides are equal and 
constant. 
A strict clock model assumes that the mutation rates 
of all lineages are equal and constant.

@@

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \texttt{model\_type} &
    \texttt{run\_if} &
    \texttt{do\_measure\_evidence} & 
    \texttt{inference model} \\ 
    \hline
    generative &
    always &
    FALSE &
    JC, strict, Yule \\
    \hline
  \end{tabular}
  \caption{
    Inference conditions and model.
    JC: Jukes-Cantor nucleotide substitution model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior.
  }
  \label{tab:RQ1}
\end{table}

@@

\begin{table}
  \begin{tabular}{ | c | c | c | l | }
    \hline
    \textbf{model\_type} &
    \textbf{run\_if} &
    \textbf{do\_measure\_evidence} & 
    \textbf{inference model} \\ 
    \hline
    generative & always         & FALSE & JC, strict, Yule \\
    candidate  & best candidate & TRUE  & JC, strict, BD   \\
    candidate  & best candidate & TRUE  & JC, strict, CBS  \\
    ...        & ...            & ...   & ...              \\
    candidate  & best candidate & TRUE  & GTR, RLN, CCP    \\
    candidate  & best candidate & TRUE  & GTR, RLN, CEP    \\
    \hline
  \end{tabular}
  \caption{
    Inference conditions and model.
    JC: Jukes-Cantor nucleotide substitution model.
    strict: strict clock model.
    Yule: Yule (pure-birth) tree prior.
    BD: birth-death tree prior.
    GTR: GTR nucleotide substitution model.
    RLN: relaxed log-normal clock model.
    CBS: coalescent Bayesian Skyline tree prior.
    CCP: coalescent constant-population tree prior.
    CEP: coalescent exponential-population tree prior.
  }
  \label{tab:RQ2}
\end{table}

@@

Here we specify a different set of experiments: we
need to state that we already have an experiment for
the generative model, as well as that we want all the other inference models 
to compete. We call the competing models 'candidate models'.
As model selection is commonly performed on the full list of available 
candidate models, \verb;pirouette; has a dedicated function 
for this choice: \verb;create_all_experiments; creates a full set 
of 40 experiments, 
containing the inference models of all combinations of 4 nucleotide substitution model, 
2 clock models and 5 tree priors. All we need to add is to exclude the 
inference model in the generative experiment:
